---
title: "Zero to Hero Bootcamp"
subtitle: "Day 1: Fundamentals of Time Series Econometrics in R"
author: "Dr Christian Engels"
date: "15 November 2024"
published-title: "Last updated"
# format: 
#   html:
#     toc: true
#     toc-location: right-body
#     toc-depth: 1
#     toc-title: "Contents"
#     html-math-method: katex
#     theme: flatly
#     number-sections: false
#     self-contained: true
#     title-block-banner: true
---

# Workshop aims

Day 1 of this workshop introduces participants to foundational concepts and techniques in time series analysis using R. It focuses on equipping attendees with practical skills for handling, analyzing and modeling time series data, with applications relevant to finance and economics.

Indicative topics:

1.  Analysis with Tidyverse
2.  Exploring Time Series Data
3.  Stationarity and Random Walks
4.  ARIMA modelling
5.  Preview of volatility processes

# Learning outcomes

By the end of the first workshop day, participants will be able to:

- Use Tidyverse tools for data analysis in R
- Explore and visualize time series data effectively
- Understand and apply concepts of random walks and stationarity
- Develop and interpret ARIMA models for forecasting
- Gain insight into volatility processes

# Where and when?

-   **Date**: 16 November
-   **Time**: 10:00-16:00
-   **Location**: WB-0001 [(Google Maps Location)](https://www.google.co.uk/maps/place/Durham+University+Business+School/@53.9879544,-4.1095165,7z/data=!4m7!3m6!1s0x487e7d000a7fdd8f:0x42b4506fe1c14d86!8m2!3d54.7803969!4d-1.5754568!15sChJ3YXRlcnNpZGUgYnVpbGRpbmdaFCISd2F0ZXJzaWRlIGJ1aWxkaW5nkgEKdW5pdmVyc2l0eZoBJENoZERTVWhOTUc5blMwVkpRMEZuU1VOSU16aHlRMnAzUlJBQuABAA!16s%2Fg%2F11y6f6m247?entry=tts&g_ep=EgoyMDI0MTExMS4wIPu8ASoASAFQAw%3D%3D)

# What to bring?

-   Your own **laptop**
-   Installation of **R and RStudio**
    - For R, follow [this link](https://www.stats.bris.ac.uk/R/)
    - For RStudio, follow [this link](https://posit.co/download/rstudio-desktop/)
    - **On Windows**, also install RTools 4.4 from [this link](https://cran.rstudio.com/bin/windows/Rtools/rtools44/rtools.html)
-   Installation of **R packages** (details below)

[Please prepare -- all download and installation steps can take 1 hour, or longer!]{style="color:red;"} If you already have an installation, please update to newest version.

After installing R, run this command:

```{r, eval=FALSE}
R.Version()
```

And verify that your R Version is the following:

```         
"R version 4.4.2 (2024-10-31)"
"Pile of Leaves"
```

In RStudio, verify that this is your version:

```         
Version 2024.09.1+394 (2024.09.1+394)
```

To install the necessary packages, run this below code in RStudio:

```{r, eval=FALSE}
posit_repo <- 
  "https://packagemanager.posit.co/cran/latest"

install.packages(
    "renv", 
    repos = posit_repo,
    ask = FALSE,
    dependencies = TRUE
  )

workshop_packages <- 
    c(
      "tidyverse",
      "tsibble",
      "feasts",
      "fable",
      "tsgarch",
      "tidyfinance",
      "httr2",
      "fixest",
      "flextable",
      "xts",
      "urca"
    )

renv::install(
    workshop_packages, 
    repos = posit_repo, 
    dependencies = "all",
    prompt = FALSE
  )
```


# Useful resources

- Textbook: "Applied Econometric Time Series (3e)" by Walter Enders
-   Online book: [Data Science for R](https://r4ds.hadley.nz)
    -   Chapter  ["Introduction"](https://r4ds.hadley.nz/intro)
    -   Chapter ["Whole Game"](https://r4ds.hadley.nz/whole-game)
-   Online book: [Tidy Finance with R](https://www.tidy-finance.org/r/)
    -   Chapter ["Accessing and Managing Financial Data"](https://www.tidy-finance.org/r/accessing-and-managing-financial-data.html)
-   Online book: [Forecasting: Principles and Practice](https://otexts.com/fpp3/)
    - Chapter [7: Time Series Regression Models](https://otexts.com/fpp3/regression.html)
    - Chapter [9: ARIMA models](https://otexts.com/fpp3/arima.html)

# Questions?

Two options:

1.  Post on [this Padlet](https://padlet.com/tsworkshop1/time-series-analysis-3mit3r31fl806b5r) (password: ts-workshop). This is the **preferred option** as there will be a lot of participants at the workshop.
2.  Contact me at [christian_engels\@outlook.com](mailto:christian_engels@outlook.com). I may not be able to respond to all requests, but I will try my best!


